### 2022-06-10 README.Rmd
readme

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron

This repository contains the R code and other materials for the paper "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron" (soon to be available as a preprint).

<!-- If you clone the repository, make sure to run the script `R/clean_fff.R` to write the files *fff_combined.csv* and *mals_combined.csv*, which have not been uploaded due to their size. When rendering, run `R/paper_build.R` to get the proper formatting for Arabic numeral citations. -->

### 2022-06-14 Rmarkdown/paper-si.Rmd
paper/data

---
title: "Supplementary information for: *Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron*"
author: "Benjamin F. Trueman, Javier Locsin, Evelyne Dor&eacute;, Kalli Hood, and Graham A. Gagnon"
date: "Centre for Water Resources Studies,* Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    keep_md: no
    reference_docx: StylesTemplate.docx
  html_document: default
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

This document contains X tables and X figures.

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper-si.Rmd")
knitr::opts_chunk$set(
  echo = FALSE,
  dev = "png",
  dpi = 600,
  fig.width = 3.33,
  fig.height = 4
)
options(dplyr.summarise.inform = FALSE)
palette <- wesanderson::wes_palette("Zissou1", 6, type = "continuous")
```

```{r packages, message=FALSE}
library("tidyverse")
library("fffprocessr")
source(here::here("R/functions.R"))
```

```{r theme}
theme_set(
  theme_minimal() + 
    theme(
      legend.position = "bottom",
      legend.margin = margin(r = 1, unit = "cm")
    )
)
```

```{r load, message=FALSE, results="hide"}

c(
  combined = "data/fff_combined.csv",
  mals_data = "data/mals_combined.csv",
  calib_icp = "data/icp_calibration_files_clean.csv",
  direct_quant = "data/icp_direct_quant.csv",
  qc_recovery = "data/qc_recoveries.csv",
  image_j = "data/image_j_clean.csv",
  proteins = "data/proteins_clean.csv",
  pdat = "data/polyp_clean.csv"
) %>% 
  map(~ read_csv(here::here(.x))) %>% 
  list2env(envir = .GlobalEnv)

```

```{r plot-names}

param_names <- tibble::tribble(
   ~param,                       ~plot_param,
  "208Pb",       "Pb (&mu;g L<sup>-1</sup>)",
    "31P",        "P (&mu;g L<sup>-1</sup>)",
   "57Fe",       "Fe (&mu;g L<sup>-1</sup>)",
   "LS90", "90&deg; light scattering signal",
"UV254_2",                "UV<sub>254</sub>",
   "ls90",         "Radius of gyration (nm)"
)

sample_names <- tibble::tribble(
          ~sample,                                                   ~plot_sample,
        "100SiO2",      "Sodium silicate (100 mg SiO<sub>2</sub> L<sup>-1</sup>)",
            "3PP",             "Sodium hexametaphosphate (3 mg P L<sup>-1</sup>)",
 "FeOx-pHcontrol",                                                      "Control",
     "100SiO2+Pb", "Sodium silicate + Pb (100 mg SiO<sub>2</sub> L<sup>-1</sup>)",
         "3PP+Pb",        "Sodium hexametaphosphate + Pb (3 mg P L<sup>-1</sup>)",
        "HM_2ppm",             "Sodium hexametaphosphate (2 mg P L<sup>-1</sup>)",
  "TrimetaP_2ppm",              "Sodium trimetaphosphate (2 mg P L<sup>-1</sup>)"
   )

```

```{r fig-s1, fig.cap="**Figure S1:** Fractograms representing pure solutions of sodium hexameta- and trimeta-phosphate.", fig.height = 2.5, fig.width = 3.33}

pdat %>%
  left_join(sample_names, by = "sample") %>% 
  filter(time > 10, sample != "blank") %>% 
  ggplot(aes(time, conc, col = plot_sample, group = interaction(date, sample))) + 
  geom_line() + 
  scale_color_manual(values = palette[c(1,4)]) +
  labs(
    x = "Retention time (min.)",
    y = expression("[P] (µg L"^-1*")"),
    col = NULL
  ) + 
  guides(col = guide_legend(nrow = 2)) +
  theme(
    legend.text = ggtext::element_markdown()
  )

```

```{r checks, eval=FALSE}

# check that peak broadening is minimal between UV and ICP:

combined %>% 
  filter(
    param %in% c("UV254_2", "57Fe"),
    sample != "FeOx-pHcontrol",
    date < "2021-12-01"
  ) %>% 
  group_by(date, sample, param) %>% 
  mutate(conc = conc / max(conc)) %>% 
  ggplot(aes(time, conc, col =  param, group = interaction(date, param))) + 
  facet_wrap(vars(sample), scales = "free_y") +
  geom_line()

```

```{r mals-plot, eval=FALSE}

# Plot the data at all 21 angles

mals_data %>% 
  filter(time > 10, date < "2021-12-01") %>% 
  ggplot(aes(time, conc, col = sample, group = interaction(date, sample))) + 
  facet_wrap(
    vars(fct_reorder(param, str_extract(param, "\\d+") %>% as.numeric())), 
    scales = "free_y"
  ) + 
  geom_line() + 
  theme(legend.position = "bottom") +
  guides(col = guide_legend(nrow = 2)) +
  labs(col = NULL)
```

### 2022-06-14 Rmarkdown/paper.Rmd
paper/data

---
title: "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron"
author: "Benjamin F. Trueman, Javier Locsin, Evelyne Dor&eacute;, Kalli Hood, and Graham A. Gagnon"
date: "Centre for Water Resources Studies,* Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  bookdown::word_document2: 
    fig_caption: true
    number_sections: false
    global_numbering: true
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(
  echo = FALSE,
  dev = "png",
  dpi = 600,
  fig.width = 3.33,
  fig.height = 4
)
options(dplyr.summarise.inform = FALSE)
palette <- wesanderson::wes_palette("Zissou1", 6, type = "continuous")
```

```{r packages, message=FALSE}
library("tidyverse")
library("fffprocessr")
source(here::here("R/functions.R"))
```

```{r theme}
theme_set(
  theme_minimal() + 
    theme(
      legend.position = "bottom",
      legend.margin = margin(r = 1, unit = "cm")
    )
)
```

```{r load, message=FALSE, results="hide"}

c(
  combined = "data/fff_combined.csv",
  mals_data = "data/mals_combined.csv",
  calib_icp = "data/icp_calibration_files_clean.csv",
  direct_quant = "data/icp_direct_quant.csv",
  qc_recovery = "data/qc_recoveries.csv",
  image_j = "data/image_j_clean.csv",
  proteins = "data/proteins_clean.csv",
  pdat = "data/polyp_clean.csv"
) %>% 
  map(~ read_csv(here::here(.x))) %>% 
  list2env(envir = .GlobalEnv)

```

```{r plot-names}

param_names <- tibble::tribble(
   ~param,                       ~plot_param,
  "208Pb",       "Pb (&mu;g L<sup>-1</sup>)",
    "31P",        "P (&mu;g L<sup>-1</sup>)",
   "57Fe",       "Fe (&mu;g L<sup>-1</sup>)",
   "LS90", "90&deg; light scattering signal",
"UV254_2",                "UV<sub>254</sub>",
   "ls90",         "Radius of gyration (nm)"
)

sample_names <- tibble::tribble(
          ~sample,                                                   ~plot_sample,
        "100SiO2",      "Sodium silicate (100 mg SiO<sub>2</sub> L<sup>-1</sup>)",
            "3PP",             "Sodium hexametaphosphate (3 mg P L<sup>-1</sup>)",
 "FeOx-pHcontrol",                                                      "Control",
     "100SiO2+Pb", "Sodium silicate + Pb (100 mg SiO<sub>2</sub> L<sup>-1</sup>)",
         "3PP+Pb",        "Sodium hexametaphosphate + Pb (3 mg P L<sup>-1</sup>)",
        "HM_2ppm",             "Sodium hexametaphosphate (2 mg P L<sup>-1</sup>)",
  "TrimetaP_2ppm",              "Sodium trimetaphosphate (2 mg P L<sup>-1</sup>)"
   )

```

# Abstract

Sequestrants such as polyphosphate and sodium silicate are used widely to control iron precipitation in drinking water, but less is known about their impact on iron corrosion scale. Here we characterize the nanoparticulate iron (oxyhydr)oxide suspensions that result from corroding cast iron coupons in solutions containing either sodium hexametaphosphate (3 mg P L^-1^) or sodium silicate (100 mg SiO~2~ L^-1^). We determined the elemental composition and size distribution of these suspensions using flow field-flow fractionation with ultraviolet, multielement, and multiangle light scattering detection (FFF-UV-MALS-ICP-MS). We used light scattering data to estimate the radius of gyration, *r~g~*, via the Zimm model, with *r^2^* ranging from 0.983--0.999. Polyphosphate and sodium silicate suspensions yielded particles in two size fractions. The smaller particles had radii of gyration equaling 21–26 nm at peak ^57^Fe intensity, and the larger particles had radii equaling 70–91 nm.  Lead associated preferentially with these particles over free hexametaphosphate, suggesting that sequestrants may interact with iron corrosion scale to yield effective transport vectors for lead in drinking water systems.

# Introduction

Polyphosphates and sodium silicates are used widely as drinking water additives, often to control aesthetic issues caused by iron from source water and pipe corrosion. These additives have been shown to decrease the colour and turbidity of iron suspensions [@robinson_effects_1987; @robinson_iron_1992; @lytle_effect_2002; @rahman_benchscale_2014; @li_understanding_2019]. They reduce particle size, alter morphology, and increase stability by imparting strongly negative zeta potentials [@magnuson_characterization_2001; @lytle_effect_2002; @rahman_benchscale_2014; @li_understanding_2019]. There is also some evidence that they control iron corrosion [@lahodny_sarc_influence_1981; @koudelka_nature_1982].

Less is known about the effect of polyphosphate and sodium silicate on the release of iron from iron corrosion scale. In the short term, an elevated dose of sodium silicate (100 mg SiO~2~ L^-1^) may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available. The effect of polyphosphate on iron colloid release from corrosion scale is also unclear. Iron colloids are important vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016].

Here, we characterize the nanoparticulate iron released from corroding cast iron coupons in the presence of sodium hexametaphosphate---a common polyphosphate additive---and sodium silicate. We determined the concentration and size distribution of colloids (1--1000 nm) and nanoparticles (1--100 nm) using flow field-flow fractionation (FFF) with ultraviolet absorbance, multielement, and multiangle light scattering detectors. We find that both sequestrants yield stable iron nanoparticle  suspensions with similar size distributions and that these colloids bind lead readily.

# Methods

## Iron suspensions

To generate each suspension, we submerged two cast iron coupons in 50 mL of solution held inside a 50 mL polypropylene tube with a polyethylene cap (coupons were 76 &times; 13 &times; 1.5 mm^3^ and sourced from BioSurface Technologies). Immediately before submersion, coupons were polished with 80 and then 400 grit SiC paper.

Each solution contained 35 mg NaHCO~3~ L^-1^ (5 mg C L^-1^, Fisher Chemical) and either sodium hexametaphosphate (3 mg P L^-1^, Alfa Aesar), sodium silicate (100 mg SiO~2~ L^-1^ with a NaO:SiO~2~ ratio of 3.22, National Silicates), or no additive (the control suspension). After preparation and mixing, solution pH was adjusted to 7.5 &plusmn; 0.2 with HNO~3~ or NaOH (Fisher Chemical).

Coupons were left to stagnate for 24 hours. Afterward, each suspension was mixed by gentle swirling. Suspensions were then filtered through 0.45 µm cellulose nitrate membrane filters (Whatman) using a syringe-mounted apparatus. To minimize adsorption, the first 10 mL of each aliquot was filtered to waste, and the next 20 mL was divided between two smaller polypropylene sample tubes; one for analysis by FFF-UV-MALS-ICP-MS and the other for direct quantification of the same elements by ICP-MS after acidification to pH < 2 with concentrated trace metal grade HNO~3~ (Fisher Chemical). To characterize partitioning of lead, we added 200 µg Pb L^-1^ as Pb(NO~3~)~2~ (Fisher Chemical) to filtered sodium silicate and hexametaphosphate suspensions. Lead was allowed to react for 53 minutes before analysis by FFF.

## Field-flow fractionation

Iron suspensions were fractionated using a modified version of the method described in a recent paper [@trueman_characterizing_2019]. We used an asymmetric flow FFF system (Postnova AF2000 Multiflow) with a 300 Da poly(ethersulfone) membrane and an autosampler with a 1 mL polyether ether ketone sample loop. The system was coupled sequentially to a UV absorbance detector (Shimadzu SPD-20A), a multiangle light scattering detector (Postnova PN3621), and an ICP-MS (ThermoFisher iCAP-RQ). The mobile phase for all separations was 50 mM tris (hydroxymethyl)aminomethane, adjusted to pH 7.3 with trace metal grade HCl. FFF channel effluent was mixed with internal standards (Sc, In, and Tb) in 2% HNO~3~ using a
mixing tee.

Method parameters are summarized in Table \@ref(tab:table-run-params). The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward it was set to decay linearly over 2 minutes to 0.1 mL min^-1^. The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; afterward, it was set to zero for the rinse step (the final 10 minutes of the run).

```{r table-run-params, echo=FALSE}
tibble::tribble(
  ~Parameter, ~Value,
  "Injection volume", "0.25 (sodium silicate) and 0.5 mL (hexametaphosphate)",
  "Spacer", "500 µm",
  "Total run time", "53 min",
  "Detector flow", "1 mL min^-1^",
  "Injection flow", "0.2 mL min^-1^",
  "Focus flow", "2.8 mL min^-1^",
  "Focus period", "10 min",
  "Transition period", "1 min",
  "Initial crossflow", "2 mL min^-1^ (first 31 min of each run)",
  "Crossflow decay", "Linear over 2 min",
  "Final crossflow", "0.1 mL min^-1^ (10 min)",
  "Rinse step", "10 min (no crossflow)"
) %>% 
  knitr::kable(caption = "FFF run parameters.")
```

## Inductively coupled plasma mass spectrometry

```{r qc-data}

to_list <- function(x) {
  x %>% 
    str_replace("(\\d+)([aA-zZ]+)", "^\\1^\\2") %>% 
    glue::glue_collapse(sep = ", ", last = " and ")
}

calib_curves <- calib_icp %>% 
  filter(param %in% c("31P", "57Fe")) %>% 
  group_by(param) %>% 
  summarize(r_sq = signif(median(r_sq), 3)) %>% 
  ungroup() %>% 
  mutate(isotope = str_extract(param, "\\d+") %>% as.numeric()) %>% 
  arrange(isotope) %>% 
  summarize(across(everything(), to_list))

qc <- qc_recovery %>% 
  filter(param %in% c("31P", "57Fe")) %>% 
  group_by(sample, param) %>% 
  summarize(value = signif(median(value), 2)) %>% 
  ungroup() %>% 
  group_by(sample) %>% 
  summarize(
    param = if_else(is.na(value), NA_character_, param),
    recovery = to_list(value),
    param_list = to_list(param)
  ) %>% 
  distinct(sample, recovery, param_list) %>% 
  pivot_wider(names_from = sample, values_from = c(recovery, param_list)) %>% 
  janitor::clean_names()

dl <- combined %>% 
  filter(param %in% c("31P", "57Fe")) %>% 
  distinct(date, param, three_sigma) %>% 
  group_by(param) %>% 
  summarize(
    three_sigma = signif(median(three_sigma), 2),
    n = n()
  ) %>% 
  ungroup() %>% 
  summarize_at(vars(c(param, three_sigma, n)), ~ to_list(unique(.x)))

```

ICP-MS data were acquired using a ThermoFisher iCAP-RQ operated in kinetic energy discrimination mode with He as the collision gas. We calibrated the ICP-MS on each analysis day using multielement standards in 2% HNO~3~ at 25, 100, and 250 µg L^-1^. Standards were introduced to the nebulizer using the ICP-MS autosampler after mixing with FFF channel effluent at the mixing tee. Calibration curves yielded median *r^2^* values of `r calib_curves$r_sq` for `r calib_curves$param`.

Quality control spikes in the same matrix (75 µg L^-1^) yielded median recoveries equaling `r qc$recovery_qc_75ppb`% of `r qc$param_list_qc_75ppb`, respectively. An independently-prepared spike at 40 µg L^-1^ yielded a median `r qc$param_list_qc_ext_40ppb` recovery of `r qc$recovery_qc_ext_40ppb`%. Detection limits were estimated using the 3$\sigma$ method, with $\sigma$ calculated separately for the elution step of `r dl$n` blanks. Median instantaneous detection limits were estimated at `r dl$three_sigma` µg L^-1^ for `r dl$param`.

The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients taken from literature [@clough_role_1981; @gaigalas_diffusion_1992].

```{r calculate-thickness,message=FALSE}

protein_peaks <- proteins %>%
  filter(time < 18, sample != "blank") %>% 
  peak_maxima(focus = 5) %>% 
  pivot_wider(id_cols = -conc, names_from = sample, values_from = time)

w_true_fer <- calculate_w(t1 = protein_peaks$ferritin, D = .35e-10, focus = 5, Vc = .0015, Vin = 2e-4) # ferritin
w_true_bsa <- calculate_w(protein_peaks$BSA, D = .632e-10, focus = 5, Vc = .0015, Vin = 2e-4) # bsa

```

## Transmission electron microscopy (TEM)

At least 13 TEM images were acquired to represent each iron suspension, using a JEOL 1230 instrument. Suspensions were vortexed, and then a 50 µL aliquot from each was loaded onto a 3 mm copper TEM grid (Formvar/Carbon FCF200). Images were scaled and then processed using ImageJ 1.53k [@schneider_nih_2012] by thresholding to differentiate particles from the background. The Feret diameter---the maximum of all transects across each perimeter---was estimated for each particle. We used a lower cutoff for particle identification of 335 nm^2^, which corresponds to the smallest spherical particle that the MALS detector can identify, based on manufacturer specifications (8 nm nominal radius of gyration).

## Data analysis

We analyzed and presented the data using R version 4.1.1 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook]. Material necessary to reproduce the results is available on GitHub [@fff_mals] and functions for post-processing FFF data are available as a separate R package [@fffprocessr].

### Peak fitting

We estimated the fraction of colloidal iron present in each of the two light scattering peaks by fitting a two-peak model to each fractogram. Each peak was represented as a skewed Gaussian and the fractogram as their sum.

### Multiangle light scattering

We estimated the radius of gyration ($r_g$) associated with each light scattering peak and fractogram using the Zimm model [@kammer_fieldflow_2005; @wyatt_light_1993; @baalousha_size_2006]. Specifically, we solved the following equation for $r_g^2$, the mean squared radius of gyration:

$$\frac{Kc}{R(\theta)} = \frac{1}{M} + \frac{\langle{r^2_g}\rangle}{3M}\left[\frac{4\pi\eta_0}{\lambda}sin(\frac{\theta}{2})\right]^2$$

where $K$ is a constant, $c$ and $M$ are the concentration and molar mass of the analyte, respectively, $\lambda$ is the wavelength of the incident light, $\theta$ is the scattering angle, $\eta_0$ is the index of refraction of the solvent, and $R$ is the Rayleigh ratio, or the scattering intensity above baseline. For environmental colloids, $K$, $c$, and $M$ are usually unknown, but $r_g^2$ can be estimated by regressing $R^{-1}$ on $sin^2(\theta/2)$. That is,

$$\langle{r^2_g}\rangle = \frac{3\beta_1\lambda^2}{16\beta_0\pi^2}$$

where $\beta_0$ and $\beta_1$ are the intercept and slope of the linear regression. Before fitting the model, data collected at the smallest and largest two scattering angles (7, 12, 156, and 164&deg;) were removed due to excessive noise (Figure \@ref(fig:figure-zimm)).

```{r calculate-rg}
mals_rg <- mals_data %>% 
  filter(
    time > 10, 
    !param %in% c("ls7", "ls12", "ls156", "ls164"),
  ) %>% 
  calculate_rg(window = .05, method = "zimm") 
```

```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999.", message=FALSE}

# calculate peak light scattering for each sample:

peak_pos <- combined %>% 
  filter(sample != "FeOx-pHcontrol", date < "2021-12-01") %>% 
  peak_maxima(focus = 33, peaks = 2) %>%
  # change peak numbering from decreasing intensity to increasing retention time:
  arrange(sample, date, param, time) %>% 
  group_by(date, sample, param) %>% 
  mutate(peak = 1:2) %>% 
  ungroup()

zimm_plot_in <- mals_rg %>% 
  left_join(
    filter(peak_pos, param == "LS90"), 
    by = c("date", "sample"), 
    suffix = c("", "_peak")
  ) %>% 
  left_join(sample_names, by = "sample") %>% 
  filter(time == time_peak) %>% 
  mutate(
    plot_sample = str_extract(plot_sample, "(?<=\\().+(?=\\))"),
    x = sin(pi * theta / 360) ^ 2,
    y = 1e-3 * 1 / rayleigh_ratio
  ) %>% 
  group_by(date, sample, peak) %>% 
  mutate(
    rsq = signif(cor(x, y) ^ 2, 3), 
    max_y = max(y),
    max_x = max(x)
  ) %>% 
  ungroup()

zimm_plot_in %>% 
  mutate(plot_sample = str_replace(plot_sample, "\\s\\(", "<br>(")) %>% 
  ggplot(aes(x, y, group = interaction(date, sample))) + 
  facet_grid(
    cols = vars(plot_sample),
    rows = vars(peak = paste0("Peak ", peak)),
    scales = "free_y"
  ) + 
  geom_smooth(method = "lm", col = palette[1]) +
  geom_point(size = 2, alpha = .5, shape = 16) +
  geom_label(
    data = function(x) distinct(x, date, sample, peak, plot_sample, rsq, max_y, max_x),
    aes(x = max_x, y = max_y, label = glue::glue("R^2~'='~{rsq}")),
    show.legend = FALSE, parse = TRUE,
    alpha = .6, label.size = 0, nudge_x = -.2, nudge_y = -2e1, size = 2,
    label.padding = unit(0.05, "lines"),
  ) +
  labs(
    x = expression(italic(sin^2*(pi * theta))),
    y = expression(italic(frac(1, "R")%*%10^3)),
    col = NULL
  ) + 
  theme(
    strip.text = ggtext::element_markdown(),
    axis.text.x = element_text(angle = 25, hjust = 1)
  ) +
  guides(col = guide_legend(ncol = 2))

```

# Results and discussion

```{r peak-positions}

pks_txt <- peak_pos %>% 
  filter(param == "57Fe") %>% 
  arrange(sample, date, time) %>% 
  group_by(date, sample) %>% 
  mutate(peak = 1:2) %>% 
  ungroup() %>% 
  group_by(peak) %>% 
  summarize(
    min = round(min(time), 1),
    max = round(max(time), 1)
  ) %>% 
  pivot_wider(names_from = peak, values_from = c(min, max))

peak_pos_p <- combined %>% 
  filter(sample == "3PP", param == "31P") %>% 
  peak_maxima()

```

Sodium silicate and hexametaphosphate suspensions yielded similar iron fractograms (Figure \@ref(fig:figure-fff)), with primary and secondary light scattering peaks. These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes. Iron was also present in a broad peak, eluting between approximately 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm. This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017]. In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which matches the peak retention of sodium hexametaphosphate in pure water (Figure S1).

```{r figure-fff, fig.cap="Fractograms representing triplicate sodium silicate and hexametaphosphate suspensions and a control suspension (no sequestrant). Theoretical hydrodynamic radius was calculated at constant crossflow, based on channel geometry and flow parameters [@baalousha_size_2006]. Light scattering and absorbance fractograms are displayed as the raw detector responses. ", fig.height=6}

plot_in <- mals_rg %>% 
  filter(param == "ls90") %>%
  pivot_longer(c(rg_zimm, conc)) %>% 
  mutate(
    value = case_when(
      name == "rg_zimm" & time < 34 ~ NA_real_,
      name == "rg_zimm" & time >= 49.7 ~ NA_real_,
      name == "rg_zimm" & time >= 34 ~ value,
      TRUE ~ value
    )
  )

combined %>% 
  filter(
    time > 10, 
    param != "208Pb",
    date < "2021-12-01"
  ) %>%  
  bind_rows(filter(plot_in, name == "rg_zimm")) %>% 
  left_join(sample_names, by = "sample") %>% 
  left_join(param_names, by = "param") %>% 
  mutate(
    value = coalesce(conc, value),
    plot_param = fct_relevel(
      plot_param,
      "Fe (&mu;g L<sup>-1</sup>)", "P (&mu;g L<sup>-1</sup>)", 
      after = 0L
    )
  ) %>% 
  ggplot(aes(time, value, group = interaction(date, sample), col = plot_sample)) + 
  facet_wrap(vars(plot_param), scales = "free_y", ncol = 1) +
  geom_line() + 
  scale_x_continuous(
    position = "top",
    sec.axis = sec_axis(
      trans = function(x) 1e9 * calculate_rh(x, w = w_true_fer, Vc = 0.002, Vin = 2e-4),
      breaks = seq(0, 10, by = 2),
      guide = guide_axis(title = "Hydrodynamic radius (nm)")
    )
  ) + 
  scale_color_manual(values = palette[c(6,1,4)]) +
  theme(
    legend.text = ggtext::element_markdown(),
    strip.text = ggtext::element_markdown()
  ) +
  labs(x = "Retention time (min)", y = NULL, col = NULL) +
  guides(col = guide_legend(nrow = 3))

```

```{r tablulate-rg}

radii <- mals_rg %>% 
  left_join(
    filter(peak_pos, param == "57Fe"), 
    by = c("date", "sample"), 
    suffix = c("", "_peak")
  ) %>% 
  group_by(date, sample, peak) %>% 
  mutate(time_matched = time[which.min(abs(time - time_peak))]) %>% 
  ungroup() %>% 
  filter(time == time_matched) %>% 
  group_by(sample, peak) %>% 
  summarize(
    time = unique(time),
    rg_zimm = unique(rg_zimm),
    d_geom = 2 * unique(rg_zimm) / sqrt(3/5),
    rep = unique(date) %>% factor() %>% as.numeric()
  ) %>% 
  ungroup()

radii_text <- radii %>%
  group_by(sample, peak) %>% 
  summarize(min = min(rg_zimm), max = max(rg_zimm)) %>% 
  pivot_wider(names_from = c(sample, peak), values_from = c(min, max)) %>% 
  summarize(across(everything(), ~ signif(.x, 2)))

```

```{r peak-integration}
source(here::here("R/peak_fits.R"))
```

The estimated radii of gyration characterizing colloids in the primary and secondary peaks ranged from `r radii_text$min_100SiO2_1`--`r radii_text$max_100SiO2_1` and `r radii_text$min_100SiO2_2`--`r radii_text$max_100SiO2_2` nm in sodium silicate suspensions, and `r radii_text$min_3PP_1`--`r radii_text$max_3PP_1` and `r radii_text$min_3PP_2`--`r radii_text$max_3PP_2` nm in hexametaphosphate suspensions (Table \@ref(tab:table-radii)). While the maximum scattering intensities were similar between the two peaks, the second one accounted for an estimated 5%---or less---of the total iron concentration. This is explained by the much larger particle sizes represented in the secondary peak, and the dependence of scattering intensity on both colloid concentration and size [@korgel_vesicle_1998; @baalousha_size_based_2006].

```{r table-radii}
radii %>% 
  arrange(sample, rep, peak) %>% 
  left_join(pcent_tot, by = c("sample", "peak", "rep")) %>% 
  transmute(
    Suspension = fct_recode(sample, "Sodium silicate" = "100SiO2", "Hexametaphosphate" = "3PP") %>% 
      unfill_vec("-"),
    Replicate = unfill_vec(rep, "-"),
    Peak = peak,
    `Retention time (min)` = signif(time, 3), 
    `Radius of gyration (nm)` = signif(rg_zimm, 3),
    `Spherical equivalent geometric diameter (nm)` = signif(d_geom, 3),
    `% total Fe` = round(pcent, 1)
  ) %>% 
  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters.")
```

### Transmission electron microscopy analysis

```{r image-j-stats}

susp_summ <- image_j %>% 
  group_by(susp) %>% 
  summarize(
    m = median(Feret), 
    q25 = quantile(Feret, .25),
    q75 = quantile(Feret, .75),
    n = n()
  ) %>% 
  mutate(across(where(is.double), ~ signif(.x, 2)))

```

We identified `r paste(susp_summ$n, collapse = " and ")` particles in TEM images representing sodium silicate and hexametaphosphate suspensions, respectively, with median Feret diameters of `r paste(susp_summ$m, collapse = " and ")` nm (Figure \@ref(fig:figure-image-j)).

Since suspensions must be evaporated on a grid for TEM analysis, aggregates present in TEM images might be different from the larger particles identified by FFF-MALS. Still, there is an approximate correspondence between the equivalent spherical diameters obtained from light scattering data and the Feret distances estimated from TEM images, especially for the primary peak at a retention time of 34--35 minutes.

```{r figure-image-j, fig.cap="Histograms representing particles identified in TEM images of sodium silicate and hexametaphosphate-stabilized iron suspensions.", fig.height=2}

image_j %>%
  mutate(susp = gsub("100Si", "100SiO2", susp)) %>% 
  left_join(sample_names, by = c("susp" = "sample")) %>% 
  ggplot(aes(
    x = Feret, 
    fill = plot_sample
  )) + 
  geom_histogram(binwidth = 5, alpha = .8) +
  scale_fill_manual(values = palette[c(2,4)]) +
  scale_colour_manual(values = palette[c(2,4)]) +
  theme(
    legend.text = ggtext::element_markdown(),
    strip.text = ggtext::element_markdown()
  ) +
  labs(
    x = "Feret diameter (nm)",
    y = "Count",
    fill = NULL
  ) + 
  guides(fill = guide_legend(nrow = 2))

```

## Analyte recovery

```{r recovery}

# integrate peaks:

fff_conc <- combined %>% 
  filter(
    time > 10, 
    param %in% c("57Fe", "31P"), 
    date < "2021-12-01"
  ) %>% 
  mutate(inj_vol = if_else(str_detect(sample, "SiO2"), 2.5e-4, 5e-4)) %>% 
  group_by(
    date, sample = str_remove_all(sample, "O2"), 
    param, inj_vol
  ) %>% 
  summarize(
    conc_ppb = integrate_peak(
      time, conc, 
      injvol = unique(inj_vol)
    )
  ) %>% 
  ungroup()

# compare to direct ICP-MS quantification:
recovery <- direct_quant %>% 
  left_join(fff_conc, by = c("date", "sample", "param")) %>% 
  filter(inj_vol != 125e-6) %>% 
  group_by(sample, param) %>% 
  mutate(percent_recovery = conc_ppb / value) %>% 
  summarize(
    n = length(na.omit(percent_recovery)),
    sd_percent_recovery = 100 * sd(percent_recovery, na.rm = TRUE),
    percent_recovery = 100 * mean(percent_recovery, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(
    sample %in% c("100Si", "3PP"),
    !(sample == "100Si" & param == "31P")
  )

recovery_text <- recovery %>%
  pivot_wider(names_from = c(sample, param), values_from = ends_with("recovery")) %>% 
  summarize(across(everything(), ~ signif(.x, 3)))
```

```{r recovery-shmp, eval=FALSE}
# recovery of SHMP @ 2 mg P/L
pdat %>% 
  filter(time > 10, sample == "HM_2ppm") %>% 
  summarize(recovery = integrate_peak(time, conc, injvol = 5e-4) / 2e3)

```

We integrated the ^31^P and ^57^Fe signals over each fractogram---excluding the focus period---to estimate the concentrations of each analyte. Then, we compared these estimates to the concentrations determined by direct quantification. Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)). While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions. And since recovery of P from a sodium hexametaphosphate standard was 111%, the partial P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1). The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane.

```{r table-recov}

recovery %>% 
  transmute(
    Suspension = fct_recode(sample, "Sodium silicate" = "100Si", "Hexametaphosphate" = "3PP") %>% 
      unfill_vec("-"),
    Analyte = str_replace(param, "(\\d+)(.+)", "^\\1^\\2"), 
    `% recovery` = signif(percent_recovery, 3),
    `Standard deviation` = signif(sd_percent_recovery, 3),
    N = n
  ) %>% 
  knitr::kable(caption = "Estimated analyte recovery, by suspension type.")

```

## Potential of colloids for contaminant transport

```{r fig-pb-in}

fig_pb_in <- combined %>% 
  filter(
    time > 10, 
    date == "2021-12-01"
  )

```

```{r corr-fe-pb, eval=FALSE}

fig_pb_in %>% 
  filter(param %in% c("208Pb", "57Fe")) %>% 
  pivot_wider(c(date, sample, time), names_from = param, values_from = conc) %>% 
  arrange(sample, time) %>% 
  fill(`208Pb`, `57Fe`, .direction = "downup") %>% 
  group_by(sample) %>% 
  summarize(cor = cor(`208Pb`, `57Fe`) ^ 2)

```

Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively. The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994], and more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015].

There was little evidence, however, of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb. This suggests that lead interacts preferentially with iron oxide colloids over free hexametaphosphate. The colloids that result from iron corrosion in the presence of a sequestrant, then, may represent an effective transport vector for lead in drinking water. The preferential partitioning of lead to iron (oxyhydr)oxide colloids was reported in another recent paper, where a much larger fraction of total lead was associated with colloidal iron than colloidal dissolved organic matter, although both groups of colloids appeared to bind lead [@trueman_characterizing_2019]. Studies of natural waters have also found that lead associated preferentially with larger colloids and is strongly correlated with colloidal iron concentrations [@tanguy_removal_2011].

```{r figure-pb, fig.cap="Fractograms representing duplicate sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^. Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=6}

fig_pb_in %>% 
  left_join(sample_names, by = "sample") %>% 
  left_join(param_names, by = "param") %>% 
  ggplot(aes(time, conc, col = plot_sample, group = interaction(date, sample))) + 
  facet_wrap(vars(plot_param), scales = "free_y", ncol = 1) + 
  geom_line() + 
  scale_color_manual(values = palette[c(1,4,3,6)]) + 
  theme(
    strip.text = ggtext::element_markdown(), 
    legend.text = ggtext::element_markdown()
  ) + 
  guides(col = guide_legend(nrow = 4)) +
  labs(x = "Retention time (min)", y = NULL, col = NULL)

```

This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may bind significant quantities of lead. And even if preliminary testing yields minimal evidence of lead complexation by the sequestrant itself, its interactions with iron may result in elevated lead concentrations. Corrosion control treatment tests, then, should account for the effects of iron corrosion when it is likely to occur in the distribution system [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @trueman_comparing_2022].

# Conclusion

Sodium silicate and sodium hexametaphosphate formed nanoparticulate suspensions of iron (oxyhydr)oxide in the presence of a corroding iron surface. The size distribution of the two suspensions was similar, and particle size estimates obtained from light scattering data agreed reasonably well with those determined by TEM image analysis. Lead partitioned preferentially to iron particles over free hexametaphosphate, suggesting that the particles represent an effective transport vector for lead, and possibly for other heavy metals. This highlights a risk associated with sequestrant use in drinking water treatment: sequestrants---even those that do not form complexes with lead---may generate colloids that provide an important mobile sink for lead, potentially increasing human exposure.

# Acknowledgments

This work was funded by Mitacs through the Mitacs Accelerate Program (Reference # IT23352) and NSERC through an Industrial Research Chair program (grant no. IRCPJ: 349838-16), a Postdoctoral Fellowship (E. Dor&eacute;), and a postgraduate scholarship (K. Hood). We acknowledge the technical support of Heather Daurie and the Electron Microscopy Core Facility at Dalhousie University.

# References

### 2022-06-14 R/clean_image_j.R

@@ -1,0 +1,18 @@
+
+library("tidyverse")
+
+# 100Si-11 and 100Si-12 are very similar, use one or other, not both
+
+rawdat <- list.files(
+  "data/image-j", 
+  pattern = ".+\\.csv", full.names = TRUE
+) %>% 
+  set_names() %>% 
+  map_dfr(read_csv, .id = "file") %>% 
+  mutate(image = regmatches(file, regexpr("(?<=image-j/)\\w+-\\w+", file, perl = TRUE))) %>% 
+  filter(image != "100Si-11") %>% 
+  separate(image, c("susp", "num"), sep = "-") 
+
+write_csv(rawdat, "data/image-j-clean.csv")
+
+

### 2022-06-14 R/functions.R

@@ -1,0 +1,8 @@
+
+# this is the opposite of fill()
+
+unfill_vec <- function(x, placeholder = "") { 
+  same <- x == dplyr::lag(x)
+  ifelse(!is.na(same) & same, placeholder, as.character(x)) 
+}
+

### 2022-06-14 R/peak_fits.R

@@ -1,0 +1,86 @@
+
+# this script is designed to be run by a call to source() from paper.Rmd
+
+# two-peak model
+
+model <- combined %>% 
+  filter(
+    sample %in% c("3PP", "100SiO2"),
+    param == "57Fe",
+    date < "2021-12-01"
+  ) %>% 
+  group_by(date, sample) %>% 
+  nest() %>% 
+  ungroup() %>% 
+  mutate(
+    sub_long  = map(data, ~ filter(.x, time > 33)),
+    # peak 2:
+    sub2  = map(data, ~ filter(.x, time > 43.5)),
+    pk2 = map(sub2, ~ deconvolve_fff(.x$time, .x$conc, h = 90, mu = 45, s = 4, g = 4,)),
+    preds_pk2 = map2(pk2, data, ~ predict(.x$model, newdata = tibble(x = .y$time))),
+    coef_pk2 = map(pk2, ~ as_tibble(t(coef(.x$model)))),
+    pk2_long = map2(data, coef_pk2, ~ with(.y, fffprocessr:::skew_gaussian(.x$time, h1, mu1, s1, g1))),
+    # peak 1:
+    sub1  = map(data, ~ filter(.x, time > 33, time < 43)),
+    pk1 = map(sub1, ~ deconvolve_fff(.x$time, .x$conc, h = 200, mu = 35, s = 5, g = 4)),
+    preds_pk1 = map2(pk1, data, ~ predict(.x$model, newdata = tibble(x = .y$time))),
+    coef_pk1 = map(pk1, ~ as_tibble(t(coef(.x$model)))),
+    pk1_long = map2(data, coef_pk1, ~ with(.y, fffprocessr:::skew_gaussian(.x$time, h1, mu1, s1, g1))),
+    linmod = pmap(
+      list(x1 = pk1_long, x2 = pk2_long, y = sub_long, z = data),
+      function(x1, x2, y, z) lm(y$conc ~ x1 + x2, data = tibble(x1 = x1[z$time > 33], x2 = x2[z$time > 33]))
+    ),
+    coef_lm = map(linmod, ~ as_tibble(t(coef(.x)))),
+    lm_fit = map(linmod, fitted),
+    peak1_fit = map2(coef_lm, pk1_long, ~ .y * .x$x1),
+    peak2_fit = map2(coef_lm, pk2_long, ~ .y * .x$x2)
+  )
+
+# % of total:
+
+pcent_tot <- model %>%
+  mutate(
[…]

### 2022-06-14 R/render_paper.R

@@ -1,0 +1,25 @@
+
+library("tidyverse")
+
+# paper:
+
+path <- "Rmarkdown/paper.Rmd"
+path_temp <- str_replace(path,"(.+/)(.+)", "\\1_\\2")
+
+read_file(path) %>% 
+  # move arabic numberal refs after punctuation:
+  str_replace_all("\\s+(\\[@.*?\\])([.,;:])?", "\\2\\1") %>% 
+  write_file(path_temp)
+
+rmarkdown::render(path_temp)
+
+file.rename(str_replace(path_temp, "\\.Rmd", "\\.docx"), str_replace(path, "\\.Rmd", "\\.docx"))
+
+unlink(path_temp)
+
+# supplement:
+
+rmarkdown::render("Rmarkdown/paper-si.Rmd")
+
+
+

### 2022-06-17 Rmarkdown/paper-si.Rmd
added session info

@@
 
+```{r table-s1}
+tibble::tribble(
+  ~Parameter, ~Value,
+  "Injection volume", "50 µL",
+  "Spacer", "500 µm",
+  "Total run time", "33.5 min",
+  "Detector flow", "1 mL min^-1^",
+  "Injection flow", "0.2 mL min^-1^",
+  "Focus flow", "1.5 mL min^-1^",
+  "Focus period", "5 min",
+  "Transition period", "1 min",
+  "Crossflow", "2 mL min^-1^ ",
+) %>% 
+  knitr::kable(caption = "**Table S1.** FFF run parameters for globular protein standards.")
+```
+
 ```{r checks, eval=FALSE}

### 2022-06-17 Rmarkdown/paper.Rmd
remove .DS_Store
added session info

@@
 
-```{r table-run-params, echo=FALSE}
+```{r table-run-params}
 tibble::tribble(
@@
 The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients taken from literature [@clough_role_1981; @gaigalas_diffusion_1992].
+The FFF method for the globular protein runs is summarized in Table S1.
 
@@
 # References
+
+```{r session-info}
+writeLines(capture.output(devtools::session_info()), "session-info.txt")
+```

### 2022-06-23 README.Rmd
edits

@@
 
-<!-- If you clone the repository, make sure to run the script `R/clean_fff.R` to write the files *fff_combined.csv* and *mals_combined.csv*, which have not been uploaded due to their size.
-When rendering, run `R/paper_build.R` to get the proper formatting for Arabic numeral citations.
--->
+<!-- When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations. -->

### 2022-06-23 Rmarkdown/paper.Rmd
edits

@@
 library("fffprocessr")
+library("ggtext")
+library("glue")
 source(here::here("R/functions.R"))
@@
 
+```{r image-j-stats-pooled}
+
+image_j_aug <- image_j %>% 
+  mutate(
+    Af = Feret ^ 2 * pi / 4, # area of circle with measured Feret diameter
+    r = Area / Af, # ASTM roundness
+    ds = sqrt(Area / pi) * 2 # diameter of circle with measured Area
+  )
+
+abs_summ <- image_j_aug %>% 
+  summarize(
+    across(
+      c(Feret, r, ds), 
+      .fns = list(med = median, lq = ~ quantile(.x, .25), uq = ~ quantile(.x, .75))
+    )
+  ) %>% 
+  mutate(across(everything(), ~ signif(.x, 2)))
+
+```
+
+```{r mals-stats, eval=FALSE}
+radii %>% 
+  group_by(peak) %>% 
+  summarize(across(c(rg_zimm, d_geom), median))
+```
+
 Sequestrants such as polyphosphate and sodium silicate are used widely to control iron precipitation in drinking water, but less is known about their impact on iron corrosion scale.
@@
 We determined the elemental composition and size distribution of these suspensions using flow field-flow fractionation with ultraviolet, multielement, and multiangle light scattering detection (FFF-UV-MALS-ICP-MS).
-We used light scattering data to estimate the radius of gyration, *r~g~*, via the Zimm model, with *r^2^* ranging from 0.983--0.999.
-Polyphosphate and sodium silicate suspensions yielded particles in two size fractions.
-The smaller particles had radii of gyration equaling 21–26 nm at peak ^57^Fe intensity, and the larger particles had radii equaling 70–91 nm.
-Lead associated preferentially with these particles over free hexametaphosphate, suggesting that sequestrants may interact with iron corrosion scale to yield effective transport vectors for lead in drinking water systems.
+Both sequestrants yielded stable iron suspensions, and the pooled median radius of gyration at peak ^57^Fe intensity was 22 nm, corresponding to a sphere-equivalent geometric diameter of 57 nm.
+The median Feret diameter and ASTM roundness was `r abs_summ$Feret_med` nm and `r abs_summ$r_med`, respectively, as determined determined by transmission electron microscopy.
+Lead associated readily with iron nanoparticles, and in the hexametaphosphate suspension it associated preferentially with iron over free hexametaphosphate.
+Sequestrants then, may interact with iron corrosion scale to yield effective transport vectors for lead in drinking water systems, even when complexation of free lead by the sequestrant is negligible.
 
@@
 They reduce particle size, alter morphology, and increase stability by imparting strongly negative zeta potentials [@magnuson_characterization_2001; @lytle_effect_2002; @rahman_benchscale_2014; @li_understanding_2019].
-There is also some evidence that they control iron corrosion [@lahodny_sarc_influence_1981; @koudelka_nature_1982].
+There is also evidence from laboratory studies that they control iron corrosion [@lahodny_sarc_influence_1981; @koudelka_nature_1982].
 
@@
 We determined the concentration and size distribution of colloids (1--1000 nm) and nanoparticles (1--100 nm) using flow field-flow fractionation (FFF) with ultraviolet absorbance, multielement, and multiangle light scattering detectors.
-We find that both sequestrants yield stable iron nanoparticle  suspensions with similar size distributions and that these colloids bind lead readily.
+We also compare these data with measurements obtained from transmission electron micrographs.
+We find that both sequestrants yield stable iron nanoparticle suspensions with similar size distributions and that these colloids associate with readily with lead.
 
@@
 Images were scaled and then processed using ImageJ 1.53k [@schneider_nih_2012] by thresholding to differentiate particles from the background.
-The Feret diameter---the maximum of all transects across each perimeter---was estimated for each particle.
+Two metrics---the Feret diameter (the maximum of all transects across each particle perimeter) and the ASTM roundness [@li_evaluation_2021] (the ratio of the observed area to the area of a circle with a diameter equal to the Feret diameter)---were estimated for each particle.
 We used a lower cutoff for particle identification of 335 nm^2^, which corresponds to the smallest spherical particle that the MALS detector can identify, based on manufacturer specifications (8 nm nominal radius of gyration).
@@
 
-We analyzed and presented the data using R version 4.1.1 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook].
+We analyzed and presented the data using R version 4.2 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook].
 Material necessary to reproduce the results is available on GitHub [@fff_mals] and functions for post-processing FFF data are available as a separate R package [@fffprocessr].
@@
 This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017].
-In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which matches the peak retention of sodium hexametaphosphate in pure water (Figure S1).
+In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S1).
 
 ```{r figure-fff, fig.cap="Fractograms representing triplicate sodium silicate and hexametaphosphate suspensions and a control suspension (no sequestrant).
-Theoretical hydrodynamic radius was calculated at constant crossflow, based on channel geometry and flow parameters [@baalousha_size_2006].
-Light scattering and absorbance fractograms are displayed as the raw detector responses.
-", fig.height=6}
+Theoretical hydrodynamic radius was calculated at constant crossflow (≤ 31 min.
+retention time), based on channel geometry and flow parameters [@baalousha_size_2006].
+Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=6}
 
@@
       trans = function(x) 1e9 * calculate_rh(x, w = w_true_fer, Vc = 0.002, Vin = 2e-4),
-      breaks = seq(0, 10, by = 2),
+      breaks = seq(0, 12, by = 2),
+      labels = c(0, 2, 4, 6, 8, 10, ">10"),
       guide = guide_axis(title = "Hydrodynamic radius (nm)")
@@
     legend.text = ggtext::element_markdown(),
-    strip.text = ggtext::element_markdown()
+    strip.text = ggtext::element_markdown(),
+    # axis.text.x = element_text(hjust = 0)
   ) +
@@
 
-### Transmission electron microscopy analysis
+### Transmission electron micrograph analysis
 
@@
 
-susp_summ <- image_j %>% 
+susp_summ <- image_j_aug %>% 
   group_by(susp) %>%
   summarize(
-    m = median(Feret), 
-    q25 = quantile(Feret, .25),
-    q75 = quantile(Feret, .75),
+    across(
+      c(Feret, r, ds), 
+      .fns = list(med = median, lq = ~ quantile(.x, .25), uq = ~ quantile(.x, .75))
+    ),
     n = n()
   ) %>% 
-  mutate(across(where(is.double), ~ signif(.x, 2)))
+  mutate(
+    across(where(is.double), ~ signif(.x, 2)),
+    diam = glue::glue("{Feret_med} ({Feret_lq}--{Feret_uq})"),
+    rnd = glue::glue("{r_med} ({r_lq}--{r_uq})"),
+  )
 
@@
 
-We identified `r paste(susp_summ$n, collapse = " and ")` particles in TEM images representing sodium silicate and hexametaphosphate suspensions, respectively, with median Feret diameters of `r paste(susp_summ$m, collapse = " and ")` nm (Figure \@ref(fig:figure-image-j)).
+We identified `r paste(susp_summ$n, collapse = " and ")` particles in TEM images representing sodium silicate and hexametaphosphate suspensions, respectively, with median Feret diameters of `r paste(susp_summ$diam, collapse = " and ")` nm (Figure \@ref(fig:figure-image-j), interquartile ranges in parentheses) and roundness values of `r paste(susp_summ$rnd, collapse = " and ")` (i.e., "sub-rounded" [@li_evaluation_2021]).
 
@@
 ```{r recovery-shmp, eval=FALSE}
-# recovery of SHMP @ 2 mg P/L
+# recovery of SHMP at 2 mg P/L
 pdat %>% 
@@
 We integrated the ^31^P and ^57^Fe signals over each fractogram---excluding the focus period---to estimate the concentrations of each analyte.
-Then, we compared these estimates to the concentrations determined by direct quantification.
+Then, we compared these estimates to the concentrations determined by direct ICP-MS quantification (i.e., no FFF).
 Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
@@
     time > 10, 
-    date == "2021-12-01"
+    date == "2021-12-01",
+    param %in% c("31P", "57Fe", "208Pb")
   )
@@
 Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively.
-The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994], and more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015].
+The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994], and more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
 
-There was little evidence, however, of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
+There was little evidence here, however, of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
 This suggests that lead interacts preferentially with iron oxide colloids over free hexametaphosphate.
 The colloids that result from iron corrosion in the presence of a sequestrant, then, may represent an effective transport vector for lead in drinking water.
-The preferential partitioning of lead to iron (oxyhydr)oxide colloids was reported in another recent paper, where a much larger fraction of total lead was associated with colloidal iron than colloidal dissolved organic matter, although both groups of colloids appeared to bind lead [@trueman_characterizing_2019].
+The preferential partitioning of lead to iron (oxyhydr)oxide colloids was reported in a recent paper, where a much larger fraction of total lead was associated with colloidal iron than colloidal dissolved organic matter, although both groups of colloids appeared to bind lead [@trueman_characterizing_2019].
+In another study, a greater fraction of dissolved lead adsorbed to iron oxide particles than precipitated with orthophosphate [@shi_lead_2020].
 Studies of natural waters have also found that lead associated preferentially with larger colloids and is strongly correlated with colloidal iron concentrations [@tanguy_removal_2011].
@@
 ```{r figure-pb, fig.cap="Fractograms representing duplicate sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^.
-Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=6}
+Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=4}
 
@@
 
-This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may bind significant quantities of lead.
-And even if preliminary testing yields minimal evidence of lead complexation by the sequestrant itself, its interactions with iron may result in elevated lead concentrations.
-Corrosion control treatment tests, then, should account for the effects of iron corrosion when it is likely to occur in the distribution system [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @trueman_comparing_2022].
+This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may bind lead.
+In a recent model distribution system study, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
+That is, while polyphosphate-based sequesration may risk elevated lead due to complexation, sodium silicate may not offer much of a solution: there is still a risk that sequestration will disperse iron particles that then transport lead.
+Testing that does not account for interactions between lead and iron may miss the true risk of elevated lead caused by sequestration.
 

### 2022-06-23 R/render_paper.R

@@ -20,6 +20,3 @@ unlink(path_temp)
 # supplement:
 
 rmarkdown::render("Rmarkdown/paper-si.Rmd")
-
-
-

### 2022-06-24 Rmarkdown/paper.Rmd
writing, new refs

@@
 We also compare these data with measurements obtained from transmission electron micrographs.
-We find that both sequestrants yield stable iron nanoparticle suspensions with similar size distributions and that these colloids associate with readily with lead.
+We find that both sequestrants yield stable iron nanoparticle suspensions with similar size distributions and that these particles associate with readily with lead.
 
@@
 
-We analyzed and presented the data using R version 4.2 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook].
+We analyzed and presented the data using R version 4.2 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook; @ggtext].
 Material necessary to reproduce the results is available on GitHub [@fff_mals] and functions for post-processing FFF data are available as a separate R package [@fffprocessr].
@@
       trans = function(x) 1e9 * calculate_rh(x, w = w_true_fer, Vc = 0.002, Vin = 2e-4),
-      breaks = seq(0, 12, by = 2),
-      labels = c(0, 2, 4, 6, 8, 10, ">10"),
+      breaks = c(seq(0, 10, by = 2), 13),
+      labels = c(seq(0, 10, by = 2), ">10"),
       guide = guide_axis(title = "Hydrodynamic radius (nm)")
@@
 While the maximum scattering intensities were similar between the two peaks, the second one accounted for an estimated 5%---or less---of the total iron concentration.
-This is explained by the much larger particle sizes represented in the secondary peak, and the dependence of scattering intensity on both colloid concentration and size [@korgel_vesicle_1998; @baalousha_size_based_2006].
+This is explained by the much larger particle sizes represented in the secondary peak, and the dependence of scattering intensity on both particles concentration and size [@korgel_vesicle_1998; @baalousha_size_based_2006].
 
@@
 In another study, a greater fraction of dissolved lead adsorbed to iron oxide particles than precipitated with orthophosphate [@shi_lead_2020].
-Studies of natural waters have also found that lead associated preferentially with larger colloids and is strongly correlated with colloidal iron concentrations [@tanguy_removal_2011].
 
@@
 This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may bind lead.
-In a recent model distribution system study, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
-That is, while polyphosphate-based sequesration may risk elevated lead due to complexation, sodium silicate may not offer much of a solution: there is still a risk that sequestration will disperse iron particles that then transport lead.
-Testing that does not account for interactions between lead and iron may miss the true risk of elevated lead caused by sequestration.
+That is, complexation of lead by polyphosphate [@holm_characterizing_1990; @holm_potential_1991; @edwards_effect_2002] is not the only risk associated with sequestrant use: even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may have an important effect on lead release.
+In a recent model distribution system study, for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
+Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.
 

### 2022-06-24 Rmarkdown/paper.Rmd
readme + eq correction

@@
 
-We estimated the radius of gyration ($r_g$) associated with each light scattering peak and fractogram using the Zimm model [@kammer_fieldflow_2005; @wyatt_light_1993; @baalousha_size_2006].
+We estimated the radius of gyration ($r_g$) of iron nanoparticles in each suspension using the Zimm model [@kammer_fieldflow_2005; @wyatt_light_1993; @baalousha_size_2006; @petteys_characterization_1998].
 Specifically, we solved the following equation for $r_g^2$, the mean squared radius of gyration:
@@
 
-$$\langle{r^2_g}\rangle = \frac{3\beta_1\lambda^2}{16\beta_0\pi^2}$$
+$$\langle{r^2_g}\rangle = \frac{3\beta_1\lambda^2}{16\beta_0\pi^2\eta_0^2}$$
 

### 2022-07-06 Rmarkdown/paper.Rmd
writing & new figure

@@
 Both sequestrants yielded stable iron suspensions, and the pooled median radius of gyration at peak ^57^Fe intensity was 22 nm, corresponding to a sphere-equivalent geometric diameter of 57 nm.
-The median Feret diameter and ASTM roundness was `r abs_summ$Feret_med` nm and `r abs_summ$r_med`, respectively, as determined determined by transmission electron microscopy.
+The median Feret diameter and ASTM roundness was `r abs_summ$Feret_med` nm and `r abs_summ$r_med`, respectively, as determined by transmission electron microscopy.
 Lead associated readily with iron nanoparticles, and in the hexametaphosphate suspension it associated preferentially with iron over free hexametaphosphate.
@@
 
-Less is known about the effect of polyphosphate and sodium silicate on the release of iron from iron corrosion scale.
+Less is known about how polyphosphates and silicates impact iron release from corrosion scale.
 In the short term, an elevated dose of sodium silicate (100 mg SiO~2~ L^-1^) may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available.
@@
 
-Here, we characterize the nanoparticulate iron released from corroding cast iron coupons in the presence of sodium hexametaphosphate---a common polyphosphate additive---and sodium silicate.
+Here, we characterize the nanoparticulate iron released from corroding cast iron coupons into solutions containing commonly-used polyphosphate- and silicate-based sequestrants.
 We determined the concentration and size distribution of colloids (1--1000 nm) and nanoparticles (1--100 nm) using flow field-flow fractionation (FFF) with ultraviolet absorbance, multielement, and multiangle light scattering detectors.
 We also compare these data with measurements obtained from transmission electron micrographs.
-We find that both sequestrants yield stable iron nanoparticle suspensions with similar size distributions and that these particles associate with readily with lead.
+We find that both sequestrants yield stable iron nanoparticle suspensions with similar size distributions and that these particles associate readily with lead.
 
@@
 
-We analyzed and presented the data using R version 4.2 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook; @ggtext].
+We analyzed and presented the data using R version 4.2 [@r] and a collection of contributed packages [@tidyverse; @rmd; @rmd_book; @rmd_cbook; @ggtext; @here].
 Material necessary to reproduce the results is available on GitHub [@fff_mals] and functions for post-processing FFF data are available as a separate R package [@fffprocessr].
@@
     show.legend = FALSE, parse = TRUE,
-    alpha = .6, label.size = 0, nudge_x = -.2, nudge_y = -2e1, size = 2,
+    alpha = .6, label.size = 0, nudge_x = -.2, nudge_y = -2e1, size = 2.5,
     label.padding = unit(0.05, "lines"),
@@
 While the maximum scattering intensities were similar between the two peaks, the second one accounted for an estimated 5%---or less---of the total iron concentration.
-This is explained by the much larger particle sizes represented in the secondary peak, and the dependence of scattering intensity on both particles concentration and size [@korgel_vesicle_1998; @baalousha_size_based_2006].
+This is explained by the much larger particle sizes represented in the secondary peak, and the dependence of scattering intensity on both particle concentration and size [@korgel_vesicle_1998; @baalousha_size_based_2006].
 
@@
 
-```{r figure-image-j, fig.cap="Histograms representing particles identified in TEM images of sodium silicate and hexametaphosphate-stabilized iron suspensions.", fig.height=2}
+```{r figure-image-j, fig.cap="(a) Histograms representing particles identified in TEM images of sodium silicate and hexametaphosphate-stabilized iron suspensions.
+(b) A TEM image representing the sodium hexametaphosphate suspension.
+(c) A TEM image representing the sodium silicate suspension."}
+knitr::include_graphics(here::here("Rmarkdown/figures/fig-tem.png"))
+```
 
+```{r image-j-panel-a, eval=FALSE}
 image_j %>%
@@
 Then, we compared these estimates to the concentrations determined by direct ICP-MS quantification (i.e., no FFF).
-Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` $\pm$ `r recovery_text$sd_percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
+Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
 While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions.
@@
 Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively.
-The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994], and more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
+The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
 
-There was little evidence here, however, of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
+But there was little evidence here of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
 This suggests that lead interacts preferentially with iron oxide colloids over free hexametaphosphate.
@@
 
-Sodium silicate and sodium hexametaphosphate formed nanoparticulate suspensions of iron (oxyhydr)oxide in the presence of a corroding iron surface.
-The size distribution of the two suspensions was similar, and particle size estimates obtained from light scattering data agreed reasonably well with those determined by TEM image analysis.
-Lead partitioned preferentially to iron particles over free hexametaphosphate, suggesting that the particles represent an effective transport vector for lead, and possibly for other heavy metals.
-This highlights a risk associated with sequestrant use in drinking water treatment: sequestrants---even those that do not form complexes with lead---may generate colloids that provide an important mobile sink for lead, potentially increasing human exposure.
+Sodium silicate and sodium hexametaphosphate solutions yielded nanoparticulate suspensions of iron (oxyhydr)oxide when exposed to a corroding iron surface.
+The size distributions of the two suspensions were similar, and particle size estimates obtained from light scattering data agreed reasonably well with those determined by TEM image analysis.
+Lead appeared to bind to iron nanoparticles, and in the sodium hexametaphosphate suspension it interacted preferentially with iron over free hexametaphosphate.
+Our findings highlight a risk associated with sequestrant use in drinking water treatment: sequestrants---even those that do not form complexes with lead---may generate colloids that provide an important mobile sink for lead, potentially increasing human exposure.
 

### 2022-07-07 Rmarkdown/paper.Rmd
new draft

@@
 Both sequestrants yielded stable iron suspensions, and the pooled median radius of gyration at peak ^57^Fe intensity was 22 nm, corresponding to a sphere-equivalent geometric diameter of 57 nm.
-The median Feret diameter and ASTM roundness was `r abs_summ$Feret_med` nm and `r abs_summ$r_med`, respectively, as determined by transmission electron microscopy.
+The median Feret diameter and ASTM roundness were `r abs_summ$Feret_med` nm and `r abs_summ$r_med`, respectively, as determined by transmission electron microscopy.
 Lead associated readily with iron nanoparticles, and in the hexametaphosphate suspension it associated preferentially with iron over free hexametaphosphate.
@@
 Method parameters are summarized in Table \@ref(tab:table-run-params).
-The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward it was set to decay linearly over 2 minutes to 0.1 mL min^-1^.
+The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward, it was set to decay linearly over 2 minutes to 0.1 mL min^-1^.
 The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; afterward, it was set to zero for the rinse step (the final 10 minutes of the run).
@@
 ```{r qc-data}
-
-to_list <- function(x) {
-  x %>% 
-    str_replace("(\\d+)([aA-zZ]+)", "^\\1^\\2") %>% 
-    glue::glue_collapse(sep = ", ", last = " and ")
-}
 
@@
 
-Quality control spikes in the same matrix (75 µg L^-1^) yielded median recoveries equaling `r qc$recovery_qc_75ppb`% of `r qc$param_list_qc_75ppb`, respectively.
+Quality control spikes in the same matrix (75 µg L^-1^) yielded median recoveries equaling `r qc$recovery_qc_75ppb`% of `r qc$param_list_qc_75ppb`.
 An independently-prepared spike at 40 µg L^-1^ yielded a median `r qc$param_list_qc_ext_40ppb` recovery of `r qc$recovery_qc_ext_40ppb`%.
@@
 
-The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients taken from literature [@clough_role_1981; @gaigalas_diffusion_1992].
+The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients drawn from literature [@clough_role_1981; @gaigalas_diffusion_1992].
 The FFF method for the globular protein runs is summarized in Table S1.
@@
 
-But there was little evidence here of lead complexation with hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
-This suggests that lead interacts preferentially with iron oxide colloids over free hexametaphosphate.
-The colloids that result from iron corrosion in the presence of a sequestrant, then, may represent an effective transport vector for lead in drinking water.
-The preferential partitioning of lead to iron (oxyhydr)oxide colloids was reported in a recent paper, where a much larger fraction of total lead was associated with colloidal iron than colloidal dissolved organic matter, although both groups of colloids appeared to bind lead [@trueman_characterizing_2019].
+But there was little evidence here of lead complexation by free hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
+The preferential interaction of lead with iron over hexametaphosphate suggests that the colloids that result from iron corrosion in the presence of a sequestrant may be a particularly effective transport vector for lead.
+Preferential partitioning of lead to iron (oxyhydr)oxide colloids was reported in a recent paper, where a much larger fraction of total lead was associated with colloidal iron than colloidal dissolved organic matter (although both appeared to bind lead) [@trueman_characterizing_2019].
 In another study, a greater fraction of dissolved lead adsorbed to iron oxide particles than precipitated with orthophosphate [@shi_lead_2020].
@@
 
-This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may bind lead.
+This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may have a substantial impact on lead release and mobility.
 That is, complexation of lead by polyphosphate [@holm_characterizing_1990; @holm_potential_1991; @edwards_effect_2002] is not the only risk associated with sequestrant use: even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may have an important effect on lead release.

### 2022-08-05 Rmarkdown/paper.Rmd
presubmission edits

@@
 ---
+
+<!-- https://doi.org/10.1016/j.carbpol.2008.03.003 -->
 
@@
 
-Sequestrants such as polyphosphate and sodium silicate are used widely to control iron precipitation in drinking water, but less is known about their impact on iron corrosion scale.
+Sequestrants such as polyphosphate and sodium silicate are used widely to control iron precipitation in drinking water, but less is known about their impacts on iron corrosion scale.
 Here we characterize the nanoparticulate iron (oxyhydr)oxide suspensions that result from corroding cast iron coupons in solutions containing either sodium hexametaphosphate (3 mg P L^-1^) or sodium silicate (100 mg SiO~2~ L^-1^).
@@
 Less is known about how polyphosphates and silicates impact iron release from corrosion scale.
-In the short term, an elevated dose of sodium silicate (100 mg SiO~2~ L^-1^) may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available.
-The effect of polyphosphate on iron colloid release from corrosion scale is also unclear.
-Iron colloids are important vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016].
+In the short term, an elevated dose of sodium silicate may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available.
+The effect of polyphosphate on iron colloid release from corrosion scale is also unclear, but it could be important: iron colloids are important vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016].
+Moreover, colloidal lead can be difficult to remove at the point of use [@lytle_lead_2020; @pan_lead_2021; @purchase_understanding_2020].
 
@@
 The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward, it was set to decay linearly over 2 minutes to 0.1 mL min^-1^.
-The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; afterward, it was set to zero for the rinse step (the final 10 minutes of the run).
+The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; then it was set to zero for the rinse step (the final 10 minutes of the run).
 
@@
 ) %>% 
-  knitr::kable(caption = "FFF run parameters.")
+  knitr::kable(caption = "FFF run parameters; since sodium silicate suspensions had much higher iron content, the injection volume was halved.")
 ```
@@
   facet_wrap(vars(plot_param), scales = "free_y", ncol = 1) +
-  geom_line() + 
+  geom_line(size = .3) + 
   scale_x_continuous(
@@
   ) + 
-  scale_color_manual(values = palette[c(6,1,4)]) +
+  scale_color_manual(values = palette[c(6,4,1)]) +
   theme(
@@
 The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane.
+In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].
 
@@
   facet_wrap(vars(plot_param), scales = "free_y", ncol = 1) + 
-  geom_line() + 
-  scale_color_manual(values = palette[c(1,4,3,6)]) + 
+  geom_line(size = .3) + 
+  scale_color_manual(values = palette[c(6,4,3,1)]) + 
   theme(
@@
 
-This transport mechanism has important implications for drinking water quality: when polyphosphate or sodium silicate is added to minimize the aesthetic impacts of iron corrosion, the iron colloids that result may have a substantial impact on lead release and mobility.
-That is, complexation of lead by polyphosphate [@holm_characterizing_1990; @holm_potential_1991; @edwards_effect_2002] is not the only risk associated with sequestrant use: even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may have an important effect on lead release.
-In a recent model distribution system study, for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
+This transport mechanism has important implications for drinking water quality.
+When polyphosphate or sodium silicate is added to minimize the aesthetic effects of iron corrosion, the iron colloids that result may increase lead mobility.
+That is, complexation of lead by polyphosphate [@holm_characterizing_1990; @holm_potential_1991; @edwards_effect_2002] is not the only risk that sequestrant use poses: depending on the relative affinity of lead for the dispersed colloids versus the free sequestrant---and the relative concentrations of the two adsorbents---the dispersion of iron may have a greater effect on lead levels.
+
+Moreover, even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may increase lead release.
+In a recent model distribution system study [@li_impact_2020], for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
 Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.

### 2022-08-10 Rmarkdown/paper-si.Rmd
edits for submission to chemrxiv

@@
 
-This document contains X tables and X figures.
+This document contains 1 table and 3 figures.
 
@@
 library("fffprocessr")
+library("ggtext")
 source(here::here("R/functions.R"))
@@
 
-param_names <- tibble::tribble(
-   ~param,                       ~plot_param,
-  "208Pb",       "Pb (&mu;g L<sup>-1</sup>)",
-    "31P",        "P (&mu;g L<sup>-1</sup>)",
-   "57Fe",       "Fe (&mu;g L<sup>-1</sup>)",
-   "LS90", "90&deg; light scattering signal",
-"UV254_2",                "UV<sub>254</sub>",
-   "ls90",         "Radius of gyration (nm)"
-)
-
 sample_names <- tibble::tribble(
@@
   "TrimetaP_2ppm",              "Sodium trimetaphosphate (2 mg P L<sup>-1</sup>)"
-   )
-
-```
-
-```{r fig-s1, fig.cap="**Figure S1:** Fractograms representing pure solutions of sodium hexameta- and trimeta-phosphate.", fig.height = 2.5, fig.width = 3.33}
-
-pdat %>%
-  left_join(sample_names, by = "sample") %>% 
-  filter(time > 10, sample != "blank") %>% 
-  ggplot(aes(time, conc, col = plot_sample, group = interaction(date, sample))) + 
-  geom_line() + 
-  scale_color_manual(values = palette[c(1,4)]) +
-  labs(
-    x = "Retention time (min.)",
-    y = expression("[P] (µg L"^-1*")"),
-    col = NULL
-  ) + 
-  guides(col = guide_legend(nrow = 2)) +
-  theme(
-    legend.text = ggtext::element_markdown()
    )
@@
 
+```{r mals-plot, fig.cap="**Figure S1.** Baseline-corrected multi-angle light scattering data (triplicate sodium hexametaphosphate suspensions)."}
+
+param_mals <- distinct(mals_data, param)$param
+striplabs <- str_replace(param_mals, "(ls)(\\d+)", "\\2&deg;")
+names(striplabs) <- param_mals
+striplabs <- as_labeller(striplabs)
+
+# Plot the data at all 21 angles
+
+plot_mals <- function(suspension) {
+  mals_data %>% 
+    filter(time > 30, sample == suspension) %>% 
+    ggplot(aes(time, conc, group = interaction(date, sample))) + 
+    facet_wrap(
+      vars(fct_reorder(param, as.numeric(str_extract(param, "\\d+")))),
+      labeller = striplabs,
+      # scales = "free_y"
+    ) + 
+    geom_line(size = .3) + 
+    theme(
+      legend.position = "bottom",
+      strip.text = ggtext::element_markdown(),
+      axis.text = element_text(size = 5)
+    ) +
+    labs(
+      x = "Retention time (min)",
+      y = "Scattering intensity above baseline",
+      col = NULL
+    )
+
+}
+
+plot_mals("3PP")
+
+```
+
+```{r mals-plot-2, fig.cap="**Figure S2.** Baseline-corrected multi-angle light scattering data (triplicate sodium silicate suspensions)."}
+plot_mals("100SiO2")
+```
+
+```{r fig-s1, fig.cap="**Figure S3:** Fractograms representing pure solutions of sodium hexameta- and trimeta-phosphate.", fig.height = 2, fig.width = 3.33}
+
+pdat %>%
+  left_join(sample_names, by = "sample") %>% 
+  filter(time > 10, sample != "blank") %>% 
+  ggplot(aes(time, conc, col = plot_sample, group = interaction(date, sample))) + 
+  geom_line() + 
+  scale_color_manual(values = palette[c(1,4)]) +
+  labs(
+    x = "Retention time (min.)",
+    y = expression("[P] (µg L"^-1*")"),
+    col = NULL
+  ) + 
+  guides(col = guide_legend(nrow = 2)) +
+  theme(
+    legend.text = ggtext::element_markdown()
+  )
+
+```
+
 ```{r checks, eval=FALSE}
@@
 ```
-
-```{r mals-plot, eval=FALSE}
-
-# Plot the data at all 21 angles
-
-mals_data %>% 
-  filter(time > 10, date < "2021-12-01") %>% 
-  ggplot(aes(time, conc, col = sample, group = interaction(date, sample))) + 
-  facet_wrap(
-    vars(fct_reorder(param, str_extract(param, "\\d+") %>% as.numeric())), 
-    scales = "free_y"
-  ) + 
-  geom_line() + 
-  theme(legend.position = "bottom") +
-  guides(col = guide_legend(nrow = 2)) +
-  labs(col = NULL)
-```

### 2022-08-10 Rmarkdown/paper.Rmd
edits for submission to chemrxiv

@@
 In the short term, an elevated dose of sodium silicate may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available.
-The effect of polyphosphate on iron colloid release from corrosion scale is also unclear, but it could be important: iron colloids are important vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016].
+The effect of polyphosphate on iron colloid release from corrosion scale is also unclear, but it could be important: iron colloids are vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016].
 Moreover, colloidal lead can be difficult to remove at the point of use [@lytle_lead_2020; @pan_lead_2021; @purchase_understanding_2020].
@@
 
+The deviation of the FFF channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients drawn from literature [@clough_role_1981; @gaigalas_diffusion_1992].
+The FFF method for the globular protein runs is summarized in Table S1.
+The MALS detector was calibrated and normalized using bovine serum albumin, polystyrene sulfonate sodium salt (63.9 kDa), and a mixture of latex particles with nominal geometric diameters of 60, 125, 350 nm.
+
 ## Inductively coupled plasma mass spectrometry
@@
 
-The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients drawn from literature [@clough_role_1981; @gaigalas_diffusion_1992].
-The FFF method for the globular protein runs is summarized in Table S1.
-
 ```{r calculate-thickness,message=FALSE}
@@
 
-where $K$ is a constant, $c$ and $M$ are the concentration and molar mass of the analyte, respectively, $\lambda$ is the wavelength of the incident light, $\theta$ is the scattering angle, $\eta_0$ is the index of refraction of the solvent, and $R$ is the Rayleigh ratio, or the scattering intensity above baseline.
+where $K$ is a constant, $c$ and $M$ are the concentration and molar mass of the analyte, respectively, $\lambda$ is the wavelength of the incident light, $\theta$ is the scattering angle, $\eta_0$ is the index of refraction of the solvent, and $R$ is the Rayleigh ratio, or the scattering intensity above baseline at each detector angle (Figures S1--2).
 For environmental colloids, $K$, $c$, and $M$ are usually unknown, but $r_g^2$ can be estimated by regressing $R^{-1}$ on $sin^2(\theta/2)$.
@@
 
-```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999.", message=FALSE}
+```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999.
+Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.", message=FALSE}
 
@@
 This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017].
-In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S1).
+In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).
 
@@
   ) %>% 
-  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters.")
+  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters. Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.")
 ```
@@
 Then, we compared these estimates to the concentrations determined by direct ICP-MS quantification (i.e., no FFF).
-Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
+Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
 While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions.

### 2022-08-11 README.Rmd
update readme

@@
 
-This repository contains the R code and other materials for the paper "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron" (soon to be available as a preprint).
-
-<!-- When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations. -->
+This repository contains the R code and other materials for the [preprint](https:doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
+When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations.

### 2022-08-11 Rmarkdown/paper.Rmd
update readme

@@
 title: "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron"
-author: "Benjamin F. Trueman, Javier Locsin, Evelyne Dor&eacute;, Kalli Hood, and Graham A. Gagnon"
-date: "Centre for Water Resources Studies,* Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
+author: "Benjamin F. Trueman*, Javier Locsin, Evelyne Dor&eacute;, Kalli Hood, and Graham A. Gagnon"
+date: "Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
 output:
@@
 
+<!-- to do: fix link in preprint metadata -->
 <!-- https://doi.org/10.1016/j.carbpol.2008.03.003 -->

### 2022-08-11 README.Rmd
update readme

@@
 
-This repository contains the R code and other materials for the [preprint](https:doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
+This repository contains the R code and other materials for the [preprint](https://doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
 When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations.

### 2022-08-18 Rmarkdown/paper-si.Rmd
preparing for submission

@@
 bibliography: references.bib
-csl: environmental-science-and-technology.csl
+csl: royal-society-of-chemistry-with-titles.csl
 ---

### 2022-08-18 Rmarkdown/paper.Rmd
preparing for submission

@@
 <!-- to do: fix link in preprint metadata -->
+<!-- omit the SS sample that makes all the other LS90 fractograms hard to see? -->
 <!-- https://doi.org/10.1016/j.carbpol.2008.03.003 -->
@@
 Sequestrants then, may interact with iron corrosion scale to yield effective transport vectors for lead in drinking water systems, even when complexation of free lead by the sequestrant is negligible.
+
+# Table of contents entry
+
+Sequestrants such as polyphosphate and sodium silicate are used to control iron precipitation in drinking water, but they may generate iron nanoparticles that increase lead mobility.
+
+```{r toc-art-save, eval=FALSE}
+
+toc_art <- fig_fff_in %>% 
+  filter(sample != "FeOx-pHcontrol") %>% 
+  ggplot(aes(time, value, col = plot_param, group = interaction(sample, date))) + 
+  geom_line(
+    data = . %>% 
+      filter(plot_param == "Radius of gyration (nm)"),
+    size = .4
+  ) + 
+  geom_line(
+    data = . %>% 
+      filter(plot_param == "Fe (&mu;g L<sup>-1</sup>)") %>% 
+      mutate(value = .3 * value),
+    size = .4
+  ) + 
+  scale_y_continuous(sec.axis = sec_axis(trans = ~ . / .3)) +
+  scale_colour_manual(values = palette[c(1, 6)]) +
+  theme(
+    legend.text = element_markdown(),
+    axis.text.y.right = element_text(colour = palette[1]),
+    axis.text.y.left = element_text(colour = palette[6]),
+    legend.box.margin = margin(),
+    legend.margin = margin()
+  ) + 
+  labs(
+    col = NULL,
+    y = NULL,
+    x = "Retention time (min)"
+  )
+
+ggsave(
+  here::here("Rmarkdown/figures/toc-art.png"), toc_art, 
+  device = "png", dpi = 600, width = 8 / 2.54, height = 4 / 2.54, 
+  bg = "white"
+)
+```
+
+```{r toc-art}
+knitr::include_graphics(here::here("Rmarkdown/figures/toc-art-combined.png"))
+```
+
+# Environmental significance statement
+
+Chemical sequestrants---mainly polyphosphate and sodium silicate compounds---improve the appearance of drinking water containing iron and manganese.
+But poyphosphates in particular have a significant drawback: they increase lead solubility.
+Here we suggest a different mechanism by which sequestrants might increase lead in drinking water: both hexametaphosphate and sodium silicate disperse (oxyhydr)oxide nanoparticles from corroding iron surfaces.
+These particles represent potential transport vectors for lead and may increase exposure via drinking water.
 
@@
 
-We estimated the radius of gyration ($r_g$) of iron nanoparticles in each suspension using the Zimm model [@kammer_fieldflow_2005; @wyatt_light_1993; @baalousha_size_2006; @petteys_characterization_1998].
+We estimated the radius of gyration ($r_g$) of iron nanoparticles in each suspension using the Zimm model [@kammer_fieldflow_2005; @wyatt_light_1993; @baalousha_size_2006; @petteys_characterization_1998; @gigault_asymmetrical_2017].
 Specifically, we solved the following equation for $r_g^2$, the mean squared radius of gyration:
@@
   labs(
-    x = expression(italic(sin^2*(pi * theta))),
-    y = expression(italic(frac(1, "R")%*%10^3)),
+    x = expression(italic(sin^2*(theta / 2))),
+    y = expression(italic(frac(1, "R"*(theta)) %*% 10^-3)),
     col = NULL
@@
 
-plot_in <- mals_rg %>% 
+plot_rg_in <- mals_rg %>% 
   filter(param == "ls90") %>%
@@
 
-combined %>% 
+fig_fff_in <- combined %>% 
   filter(
@@
   ) %>%  
-  bind_rows(filter(plot_in, name == "rg_zimm")) %>% 
+  bind_rows(filter(plot_rg_in, name == "rg_zimm")) %>% 
   left_join(sample_names, by = "sample") %>% 
@@
     )
-  ) %>% 
+  )
+
+fig_fff_in %>% 
   ggplot(aes(time, value, group = interaction(date, sample), col = plot_sample)) + 

### 2022-08-19 Rmarkdown/paper.Rmd
journal submission

@@
 
-<!-- to do: fix link in preprint metadata -->
-<!-- omit the SS sample that makes all the other LS90 fractograms hard to see? -->
 <!-- https://doi.org/10.1016/j.carbpol.2008.03.003 -->
@@
 Here we characterize the nanoparticulate iron (oxyhydr)oxide suspensions that result from corroding cast iron coupons in solutions containing either sodium hexametaphosphate (3 mg P L^-1^) or sodium silicate (100 mg SiO~2~ L^-1^).
-We determined the elemental composition and size distribution of these suspensions using flow field-flow fractionation with ultraviolet, multielement, and multiangle light scattering detection (FFF-UV-MALS-ICP-MS).
+We determined the elemental composition and size distribution of these suspensions using flow field-flow fractionation with ultraviolet, multielement, and multiangle light scattering detection.
 Both sequestrants yielded stable iron suspensions, and the pooled median radius of gyration at peak ^57^Fe intensity was 22 nm, corresponding to a sphere-equivalent geometric diameter of 57 nm.
@@
 Suspensions were then filtered through 0.45 µm cellulose nitrate membrane filters (Whatman) using a syringe-mounted apparatus.
-To minimize adsorption, the first 10 mL of each aliquot was filtered to waste, and the next 20 mL was divided between two smaller polypropylene sample tubes; one for analysis by FFF-UV-MALS-ICP-MS and the other for direct quantification of the same elements by ICP-MS after acidification to pH < 2 with concentrated trace metal grade HNO~3~ (Fisher Chemical).
+To minimize adsorption, the first 10 mL of each aliquot was filtered to waste, and the next 20 mL was divided between two smaller polypropylene sample tubes; one for analysis by FFF and the other for direct quantification of the same elements by ICP-MS after acidification to pH < 2 with concentrated trace metal grade HNO~3~ (Fisher Chemical).
 To characterize partitioning of lead, we added 200 µg Pb L^-1^ as Pb(NO~3~)~2~ (Fisher Chemical) to filtered sodium silicate and hexametaphosphate suspensions.
@@
 
-```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999.
-Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.", message=FALSE}
+```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999 (row labels: *t*~r~ denotes retention time).", message=FALSE}
 
@@
     cols = vars(plot_sample),
-    rows = vars(peak = paste0("Peak ", peak)),
-    scales = "free_y"
+    rows = vars(peak),
+    scales = "free_y",
+    labeller = labeller(.rows = c(
+      "1" = "*t*<sub>r</sub> = 34&ndash;35 min", 
+      "2" = "*t*<sub>r</sub> = 44&ndash;45 min"
+    ))
   ) + 
@@
   theme(
-    strip.text = ggtext::element_markdown(),
+    strip.text.x = ggtext::element_markdown(),
+    strip.text.y = ggtext::element_markdown(),
     axis.text.x = element_text(angle = 25, hjust = 1)
@@
     `Spherical equivalent geometric diameter (nm)` = signif(d_geom, 3),
-    `% total Fe` = round(pcent, 1)
+    `% total Fe (approx.)` = round(pcent, 1)
   ) %>% 
@@
 Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively.
-The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @xie_chapter_2007; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
+The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
 

### 2022-08-19 README.Rmd
update readme

--- 2022-08-11 README.Rmd
@@
 
-This repository contains the R code and other materials for the [preprint](https:doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
+This repository contains the R code and other materials for the [preprint](https://doi.org/10.26434/chemrxiv-2022-rtzqm-v2) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
 When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations.

### 2022-09-14 Rmarkdown/paper.Rmd
edits after peer review (first batch)

@@
       filter(plot_param == "Radius of gyration (nm)"),
-    size = .4
+    size = .2
   ) + 
@@
       mutate(value = .3 * value),
-    size = .4
+    size = .2
   ) + 
@@
 
-```{r calculate-thickness,message=FALSE}
+```{r calculate-thickness, message=FALSE}
 
@@
   summarize(
-    min = round(min(time), 1),
-    max = round(max(time), 1)
+    min = round(min(time)),
+    max = round(max(time))
   ) %>% 
@@
 This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017].
-In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).
+In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at approximately `r unique(round(peak_pos_p$time))` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).
+
+The control suspension---prepared without either sodium silicate or hexametaphosphate---did not contain iron colloids that were detectable by FFF-ICP-MS (Figure \@ref(fig:figure-fff)).
+This is consistent with previous research showing that, in the absence of a stabilizing agent---polyphosphate, sodium silicate, natural organic matter, etc...---iron oxide nanoparticles tend to form aggregates with diameters greater than 450 nm, the upper size limit considered here [@zhang_impact_2009; @mylon_influence_2004; @rushing_effects_2003].
 
@@
 retention time), based on channel geometry and flow parameters [@baalousha_size_2006].
-Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=6}
+Light scattering and absorbance fractograms are displayed with the raw detector responses on the y-axis.", fig.height=6}
 
@@
     legend.text = ggtext::element_markdown(),
-    strip.text = ggtext::element_markdown(),
-    # axis.text.x = element_text(hjust = 0)
+    strip.text = ggtext::element_markdown()
   ) +
@@
   ) %>% 
-  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters. Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.")
+  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters of iron colloids.
+Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.
+(N.B., the control suspension did not contain iron colloids detectable by FFF-ICP-MS.)")
 ```
@@
 Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)).
+
 While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions.
@@
   ) %>% 
-  knitr::kable(caption = "Estimated analyte recovery, by suspension type.")
+  knitr::kable(caption = "Estimated analyte recovery, by suspension type. Since no peaks were identified in fractograms representing the control suspension, recovery of iron was approximately 0%.")
 
@@
 The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
+Given that iron colloids were not detectable in the control suspension, it was excluded from the partitioning experiment.
+The larger aggregates it contained, though, would also be expected to adsorb lead [@mckenzie_adsorption_1980].
 
-But there was little evidence here of lead complexation by free hexametaphosphate; the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
+There was little evidence here of lead complexation by free hexametaphosphate: the primary ^31^P signal at a retention time of approximately 14 minutes did not co-occur with ^208^Pb.
 The preferential interaction of lead with iron over hexametaphosphate suggests that the colloids that result from iron corrosion in the presence of a sequestrant may be a particularly effective transport vector for lead.
@@
 ```{r figure-pb, fig.cap="Fractograms representing duplicate sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^.
-Light scattering and absorbance fractograms are displayed as the raw detector responses.", fig.height=4}
+Light scattering and absorbance fractograms are displayed with the raw detector responses on the y-axis.", fig.height=4}
 
@@
 
-This transport mechanism has important implications for drinking water quality.
-When polyphosphate or sodium silicate is added to minimize the aesthetic effects of iron corrosion, the iron colloids that result may increase lead mobility.
+The interaction among corroding iron, sequestrant additives, and lead has important implications for drinking water quality.
+When polyphosphate or sodium silicate is used to minimize the aesthetic effects of iron corrosion, the iron colloids that result may increase lead mobility.
 That is, complexation of lead by polyphosphate [@holm_characterizing_1990; @holm_potential_1991; @edwards_effect_2002] is not the only risk that sequestrant use poses: depending on the relative affinity of lead for the dispersed colloids versus the free sequestrant---and the relative concentrations of the two adsorbents---the dispersion of iron may have a greater effect on lead levels.
@@
 In a recent model distribution system study [@li_impact_2020], for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
+In the long term, silicate treatment may decrease iron release even if it initially does the opposite [@rushing_effects_2003], but anticipating and preparing for the short-term implications highlighted here will help protect water quality.
 Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.
@@
 ```{r session-info}
-writeLines(capture.output(devtools::session_info()), "session-info.txt")
+writeLines(capture.output(devtools::session_info()), here::here("session-info.txt"))
 ```

### 2022-09-16 Rmarkdown/paper.Rmd
edits after peer review (2nd batch)

@@
 Suspensions were then filtered through 0.45 µm cellulose nitrate membrane filters (Whatman) using a syringe-mounted apparatus.
-To minimize adsorption, the first 10 mL of each aliquot was filtered to waste, and the next 20 mL was divided between two smaller polypropylene sample tubes; one for analysis by FFF and the other for direct quantification of the same elements by ICP-MS after acidification to pH < 2 with concentrated trace metal grade HNO~3~ (Fisher Chemical).
+To limit adsorption, the first 10 mL of each aliquot was filtered to waste, and the next 20 mL was divided between two smaller polypropylene sample tubes; one for analysis by FFF and the other for direct quantification of the same elements by ICP-MS after acidification to pH < 2 with concentrated trace metal grade HNO~3~ (Fisher Chemical).
 To characterize partitioning of lead, we added 200 µg Pb L^-1^ as Pb(NO~3~)~2~ (Fisher Chemical) to filtered sodium silicate and hexametaphosphate suspensions.
@@
   ) %>% 
-  knitr::kable(caption = "Estimated analyte recovery, by suspension type. Since no peaks were identified in fractograms representing the control suspension, recovery of iron was approximately 0%.")
+  knitr::kable(caption = "Estimated analyte recovery, by suspension type.
+Since no peaks were identified in fractograms representing the control suspension, recovery of iron by FFF was approximately 0%.")
 
@@
 The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
-Given that iron colloids were not detectable in the control suspension, it was excluded from the partitioning experiment.
+Given that iron colloids with diameters smaller than 450 nm were not detectable in the control suspension, it was excluded from the partitioning experiment.
 The larger aggregates it contained, though, would also be expected to adsorb lead [@mckenzie_adsorption_1980].

### 2022-09-16 R/functions.R

@@ -1,4 +1,6 @@
 
+# this script is called by source() from paper.Rmd, after loading tidyverse
+
 # this is the opposite of fill()
 
 unfill_vec <- function(x, placeholder = "") { 
@@ -11,5 +13,49 @@ to_list <- function(x) {
     str_replace("(\\d+)([aA-zZ]+)", "^\\1^\\2") %>% 
     unique() %>% 
     glue::glue_collapse(sep = ", ", last = " and ")
+}
+
+# calculate number of matches to a regex in references, as a percentage of total references:
+
+percent_bib_matches <- function(regex) {
+  
+  bib <- read_lines(here::here("Rmarkdown/references.bib")) %>% 
+    tibble(bib = .)
+  rmd <- read_file(here::here("Rmarkdown/paper.Rmd"))
+  
+  keys <- rmd %>% 
+    str_extract_all("(?<=[^\\\\]@)[^;\\]]+") %>% 
+    unlist() %>% 
+    tibble(key = .) %>% 
+    filter(!str_detect(key, "dal.ca")) %>% # email address
+    distinct()
+  
+  filter_bib <- function(x) {
+    bib %>% 
+      filter(
+        str_detect(
+          bib, 
+          glue::glue("@[aA-zZ]+\\{`x`,|author", .open = "`", .close = "`")
+        )
+      ) %>% 
+      filter(str_detect(lag(bib), "@[aA-zZ]"))
+  }
+  
+  sc <- keys %>% 
+    mutate(
+      extr = map(
+        key, 
+        filter_bib
+      )
+    ) %>% 
+    unnest(extr) %>% 
+    distinct() %>% 
+    pull(bib) %>% 
+    str_detect(regex) %>%
+    mean() %>% 
+    round(2)
+  
+  paste0(100 * sc, "% of references matching")
+  
 }
 

### 2022-09-20 Rmarkdown/paper.Rmd
resubmission to es-adv

@@
 The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward, it was set to decay linearly over 2 minutes to 0.1 mL min^-1^.
-The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; then it was set to zero for the rinse step (the final 10 minutes of the run).
+The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; afterward it was set to zero for the rinse step (the final 10 minutes of the run).
 
@@
 These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes.
-Iron was also present in a broad peak, eluting between approximately 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm.
+Iron was also present in a broad peak, eluting between 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm.
 This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017].
-In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at approximately `r unique(round(peak_pos_p$time))` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).
+In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at approximately `r unique(round(peak_pos_p$time))` minutes, which matches the peak retention of sodium hexametaphosphate in pure water (Figure S3).
 
-The control suspension---prepared without either sodium silicate or hexametaphosphate---did not contain iron colloids that were detectable by FFF-ICP-MS (Figure \@ref(fig:figure-fff)).
-This is consistent with previous research showing that, in the absence of a stabilizing agent---polyphosphate, sodium silicate, natural organic matter, etc...---iron oxide nanoparticles tend to form aggregates with diameters greater than 450 nm, the upper size limit considered here [@zhang_impact_2009; @mylon_influence_2004; @rushing_effects_2003].
+The control suspension---prepared without sodium silicate or hexametaphosphate---did not contain iron colloids that were detectable by FFF-ICP-MS (Figure \@ref(fig:figure-fff)).
+This aligns with research showing that, in the absence of a stabilizing agent---a sequestrant, for instance---iron oxide nanoparticles tend to form aggregates with diameters greater than 450 nm, the upper size limit considered here [@zhang_impact_2009; @mylon_influence_2004; @rushing_effects_2003].
 
@@
 
-While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions.
-And since recovery of P from a sodium hexametaphosphate standard was 111%, the partial P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1).
-The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane.
+While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by polyphosphate species may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions.
+And since recovery of ^31^P from a sodium hexametaphosphate standard was 111%, the partial ^31^P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1).
+The partial recovery of iron may be due to chelation by trimetaphosphate in particular [@rasmussen_phosphate_1986]: the resulting complex would also escape retention by the FFF membrane.
 In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].
@@
 Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively.
-The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
+The strong apparent relationship between iron and lead is consistent with research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010].
 Given that iron colloids with diameters smaller than 450 nm were not detectable in the control suspension, it was excluded from the partitioning experiment.
@@
 
-```{r figure-pb, fig.cap="Fractograms representing duplicate sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^.
+```{r figure-pb, fig.cap="Fractograms representing sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^.
 Light scattering and absorbance fractograms are displayed with the raw detector responses on the y-axis.", fig.height=4}
@@
 
-Moreover, even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may increase lead release.
+Moreover, even sequestrants that do not form strong complexes with lead may increase lead release.
 In a recent model distribution system study [@li_impact_2020], for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream.
-In the long term, silicate treatment may decrease iron release even if it initially does the opposite [@rushing_effects_2003], but anticipating and preparing for the short-term implications highlighted here will help protect water quality.
+In the long term, silicate treatment may decrease iron release even if it initially does the opposite [@rushing_effects_2003], but anticipating the short-term effects highlighted here will help protect water quality.
 Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.
@@
 Our findings highlight a risk associated with sequestrant use in drinking water treatment: sequestrants---even those that do not form complexes with lead---may generate colloids that provide an important mobile sink for lead, potentially increasing human exposure.
+
+# Author contributions
+
+Benjamin Trueman: investigation, data curation, formal analysis, visualization, writing--original draft; Javier Locsin: investigation, methodology, writing--review & editing; Evelyne Dor&eacute;: investigation, methodology, writing--review & editing; Kalli Hood: investigation, methodology, writing--review & editing; Graham Gagnon: funding acquisition, writing--review & editing.
+
+# Conflicts of interest
+
+There are no conflicts to declare.
 

### 2022-10-20 README.Rmd
update readme with link to published article

--- 2022-08-11 README.Rmd
@@
 
-This repository contains the R code and other materials for the [preprint](https:doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
+This repository contains the R code and other materials for the [paper](https://doi.org/10.1039/D2VA00199C) (and [preprint](https://doi.org/10.26434/chemrxiv-2022-rtzqm-v2)) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
 When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations.

### 2022-10-20 README.Rmd
readme link update

--- 2022-08-11 README.Rmd
@@
 
-This repository contains the R code and other materials for the [preprint](https://doi.org/10.26434/chemrxiv-2022-rtzqm) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
+This repository contains the R code and other materials for the [paper](https://doi.org/10.1039/D2VA00199C) (and [preprint](https://doi.org/10.26434/chemrxiv-2022-rtzqm-v3)) "Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron".
 When rendering the .Rmd file, run `R/render_paper.R` to get the proper formatting for Arabic numeral citations.
